---
title: "Assignment 1: Census Data Quality for Policy Decisions"
subtitle: "Evaluating Data Reliability for Algorithmic Decision-Making"
author: "Xinyuan(Christine) Cui"
date: Sep.19
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
execute:
  warning: false
  message: false
---

# Assignment Overview

## Scenario

You are a data analyst for the **Texas Department of Human Services**. The department is considering implementing an algorithmic system to identify communities that should receive priority for social service funding and outreach programs. Your supervisor has asked you to evaluate the quality and reliability of available census data to inform this decision.

Drawing on our Week 2 discussion of algorithmic bias, you need to assess not just what the data shows, but how reliable it is and what communities might be affected by data quality issues.

## Learning Objectives

- Apply dplyr functions to real census data for policy analysis
- Evaluate data quality using margins of error 
- Connect technical analysis to algorithmic decision-making
- Identify potential equity implications of data reliability issues
- Create professional documentation for policy stakeholders

## Submission Instructions

**Submit by posting your updated portfolio link on Canvas.** Your assignment should be accessible at `your-portfolio-url/assignments/assignment_1/`

Make sure to update your `_quarto.yml` navigation to include this assignment under an "Assignments" menu.

# Part 1: Portfolio Integration

Create this assignment in your portfolio repository under an `assignments/assignment_1/` folder structure. Update your navigation menu to include:

```
- text: Assignments
  menu:
    - href: assignments/assignment_1/your_file_name.qmd
      text: "Assignment 1: Census Data Exploration"
```
If there is a special character like comma, you need use double quote mark so that the quarto can identify this as text

# Setup

```{r setup}
# Load required packages (hint: you need tidycensus, tidyverse, and knitr)
"knitr" %in% rownames(installed.packages())
library(tidycensus)
library(tidyverse)
library(knitr)
# Set your Census API key
census_api_key("20068788c6e79d5716fbceb0dcd562ab23f74ca1", overwrite=TRUE, install = TRUE)
# Choose your state for analysis - assign it to a variable called my_state
my_state <- "TX"
```

**State Selection:** I have chosen **Texas** for this analysis because: Texas is an ideal state for this analysis due to its unique combination of rapid population growth, demographic diversity, and significant geographic variation. These factors present specific data quality challenges that are critical to evaluating algorithmic fairness.

# Part 2: County-Level Resource Assessment

## 2.1 Data Retrieval

**Your Task:** Use `get_acs()` to retrieve county-level data for your chosen state.

**Requirements:**
- Geography: county level
- Variables: median household income (B19013_001) and total population (B01003_001)  
- Year: 2022
- Survey: acs5
- Output format: wide

**Hint:** Remember to give your variables descriptive names using the `variables = c(name = "code")` syntax.

```{r county-data}
# Write your get_acs() code here
county_data_texas <- get_acs(
  geography = "county",        # 检索县级数据
  variables = c(                # 指定要获取的变量及其描述性名称
    median_household_income = "B19013_001",
    total_population = "B01003_001"
  ),
  state = my_state,            # 指定州
  year = 2022,                 # 指定年份
  survey = "acs5",             # 指定使用5年期调查
  output = "wide"              # 指定输出为宽格式
)

# Clean the county names to remove state name and "County" 
# Hint: use mutate() with str_remove()
county_data_texas <- county_data_texas %>%
  mutate(
    # 先移除州名（例如 ", Texas"）
    NAME = str_remove(NAME, ", Texas"),
    # 再移除 " County"
    NAME = str_remove(NAME, " County")
  )
# Display the first few rows
head(county_data_texas)
```

## 2.2 Data Quality Assessment

**Your Task:** Calculate margin of error percentages and create reliability categories.

**Requirements:**
- Calculate MOE percentage: (margin of error / estimate) * 100
- Create reliability categories:
  - High Confidence: MOE < 5%
  - Moderate Confidence: MOE 5-10%  
  - Low Confidence: MOE > 10%
- Create a flag for unreliable estimates (MOE > 10%)

**Hint:** Use `mutate()` with `case_when()` for the categories.

```{r income-reliability}
# Calculate MOE percentage and reliability categories using mutate()
county_data_texas <- county_data_texas %>%
  mutate(
    # 计算家庭收入的 MOE 百分比
    income_moe_pct = (median_household_incomeM / median_household_incomeE) * 100,
    
    # 根据百分比创建可靠性类别
    income_reliability = case_when(
      income_moe_pct < 5 ~ "High Confidence",
      income_moe_pct >= 5 & income_moe_pct <= 10 ~ "Moderate Confidence",
      income_moe_pct > 10 ~ "Low Confidence"
    ),
    
    # 创建一个不可靠的标记（如果 MOE > 10%）
    unreliable_income = income_moe_pct > 10
  )
# Create a summary showing count of counties in each reliability category
# Hint: use count() and mutate() to add percentages
county_data_texas %>%
  count(income_reliability) %>%
  mutate(
    percentage = (n / sum(n)) * 100
  )
```

## 2.3 High Uncertainty Counties

**Your Task:** Identify the 5 counties with the highest MOE percentages.

**Requirements:**
- Sort by MOE percentage (highest first)
- Select the top 5 counties
- Display: county name, median income, margin of error, MOE percentage, reliability category
- Format as a professional table using `kable()`

**Hint:** Use `arrange()`, `slice()`, and `select()` functions.

```{r high-uncertainty}
# Create table of top 5 counties by MOE percentage
library(knitr)
county_data_texas %>%
  arrange(desc(income_moe_pct)) %>%
  slice(1:5) %>%
  select(
    `County Name` = NAME,
    `Median Income` = median_household_incomeE,
    `Margin of Error` = median_household_incomeM,
    `MOE Percentage` = income_moe_pct,
    `Reliability Category` = income_reliability
  ) %>%

# Format as table with kable() - include appropriate column names and caption
kable(
    caption = "Top 5 Counties with The Highest MOE Percentages"
  )
```

**Data Quality Commentary:**

**[Write 2-3 sentences explaining what these results mean for algorithmic decision-making. Consider: Which counties might be poorly served by algorithms that rely on this income data? What factors might contribute to higher uncertainty?]**

# Part 3: Neighborhood-Level Analysis

## 3.1 Focus Area Selection

**Your Task:** Select 2-3 counties from your reliability analysis for detailed tract-level study.

**Strategy:** Choose counties that represent different reliability levels (e.g., 1 high confidence, 1 moderate, 1 low confidence) to compare how data quality varies.

```{r select-counties}
# Use filter() to select 2-3 counties from your county_reliability data
# Store the selected counties in a variable called selected_counties
selected_counties <- county_data_texas %>%
  filter(
    income_moe_pct > 30 |
    (income_moe_pct >= 5 & income_moe_pct <= 10) |
    income_moe_pct < 2
  ) %>%
  # 进一步精简，确保只选择3个
  group_by(income_reliability) %>%
  slice(1)
# Display the selected counties with their key characteristics
# Show: county name, median income, MOE percentage, reliability category
selected_counties %>%
  select(
    `County Name` = NAME,
    `Median Income` = median_household_incomeE,
    `MOE Percentage` = income_moe_pct,
    `Reliability Category` = income_reliability
  ) %>%
  kable(
    caption = "3 Counties Selected for Tract-Level Study"
  )
```

**Comment on the output:** [write something :)]
在high和low中选择了范围内较为极端的数据，意在使结果更典型，更好做对比。

## 3.2 Tract-Level Demographics

**Your Task:** Get demographic data for census tracts in your selected counties.

**Requirements:**
- Geography: tract level
- Variables: white alone (B03002_003), Black/African American (B03002_004), Hispanic/Latino (B03002_012), total population (B03002_001)
- Use the same state and year as before
- Output format: wide

**Challenge:** You'll need county codes, not names. Look at the GEOID patterns in your county data for hints.

```{r tract-demographics}
# Define your race/ethnicity variables with descriptive names
ethnicity_variables <- c(
  white_alone = "B03002_003",
  Black_African_American = "B03002_004",
  Hispanic_Latino = "B03002_012",
  total_population = "B03002_001"
)

# Use get_acs() to retrieve tract-level data
# Hint: You may need to specify county codes in the county parameter
selected_county_geoids <- selected_counties$GEOID
selected_county_codes <- str_sub(selected_county_geoids, start = -3) #the last 3 numbers of GEOID is county codes

tract_data_texas <- get_acs(
  geography = "tract",        # 普查区级别
  variables = ethnicity_variables,         # 指定要获取的变量及其描述性名称
  state = my_state,            # 指定州
  county = selected_county_codes, # 将县的 GEOID 作为 county 参数
  year = 2022,                 # 指定年份
  survey = "acs5",             # 指定使用5年期调查
  output = "wide"              # 指定输出为宽格式
)

# Calculate percentage of each group using mutate()
# Create percentages for white, Black, and Hispanic populations
tract_data_texas <- tract_data_texas %>%
  mutate(
    pct_white = (white_aloneE / total_populationE) * 100,
    pct_black = (Black_African_AmericanE / total_populationE) * 100,
    pct_hispanic = (Hispanic_LatinoE / total_populationE) * 100
  )

# Add readable tract and county name columns using str_extract() or similar
tract_data_texas <- tract_data_texas %>%
  mutate(
    # 使用 str_remove() 移除多余的普查区信息，只保留县名和州名
    NAME = str_remove(NAME, "Census Tract \\d+\\.?\\d*, "),
    
    # 移除州名，只保留县名
    NAME = str_remove(NAME, ", Texas"),
    
    # 将清理后的 NAME 列重命名为 county_name 以便更好地理解
    county_name = NAME,
    
    # 普查区编号通常保存在 GEOID 的后六位
    tract_id = str_sub(GEOID, -6)
  )

```

## 3.3 Demographic Analysis

**Your Task:** Analyze the demographic patterns in your selected areas.

```{r demographic-analysis}
# Find the tract with the highest percentage of Hispanic/Latino residents
# Hint: use arrange() and slice() to get the top tract
tract_data_texas %>%
  arrange(desc(pct_hispanic)) %>%
  slice(1) %>%
  select(
    `Tract ID` = tract_id,
    `County Name` = county_name,
    `Hispanic/Latino Percentage` = pct_hispanic,
    `Total Population` = total_populationE
  ) %>%
  kable(
    caption = "Tract with the Highest Percentage of Hispanic/Latino Residents"
  )

# Calculate average demographics by county using group_by() and summarize()
# Show: number of tracts, average percentage for each racial/ethnic group
county_demographics_summary <- tract_data_texas %>%
  # 1. 按县名进行分组
  group_by(county_name) %>%
  
  # 2. 对每个县组进行汇总计算
  summarize(
    number_of_tracts = n(), # 计算每个县的普查区数量
    avg_white_pct = mean(pct_white, na.rm = TRUE), # 计算平均白人人口百分比
    avg_black_pct = mean(pct_black, na.rm = TRUE), # 计算平均非裔人口百分比
    avg_hispanic_pct = mean(pct_hispanic, na.rm = TRUE) # 计算平均拉丁裔人口百分比
  )

# Create a nicely formatted table of your results using kable()
county_demographics_summary %>%
  kable(
    caption = "Average Demographics by County"
  )
```

# Part 4: Comprehensive Data Quality Evaluation

## 4.1 MOE Analysis for Demographic Variables

**Your Task:** Examine margins of error for demographic variables to see if some communities have less reliable data.

**Requirements:**
- Calculate MOE percentages for each demographic variable
- Flag tracts where any demographic variable has MOE > 15%
- Create summary statistics

```{r demographic-moe}
# Calculate MOE percentages for white, Black, and Hispanic variables
# Hint: use the same formula as before (margin/estimate * 100)

# Create a flag for tracts with high MOE on any demographic variable
# Use logical operators (| for OR) in an ifelse() statement

# Create summary statistics showing how many tracts have data quality issues
```

## 4.2 Pattern Analysis

**Your Task:** Investigate whether data quality problems are randomly distributed or concentrated in certain types of communities.

```{r pattern-analysis}
# Group tracts by whether they have high MOE issues
# Calculate average characteristics for each group:
# - population size, demographic percentages

# Use group_by() and summarize() to create this comparison
# Create a professional table showing the patterns
```

**Pattern Analysis:** [Describe any patterns you observe. Do certain types of communities have less reliable data? What might explain this?]


# Part 5: Policy Recommendations

## 5.1 Analysis Integration and Professional Summary

**Your Task:** Write an executive summary that integrates findings from all four analyses.

**Executive Summary Requirements:**
1. **Overall Pattern Identification**: What are the systematic patterns across all your analyses?
2. **Equity Assessment**: Which communities face the greatest risk of algorithmic bias based on your findings?
3. **Root Cause Analysis**: What underlying factors drive both data quality issues and bias risk?
4. **Strategic Recommendations**: What should the Department implement to address these systematic issues?

**Executive Summary:**

[Your integrated 4-paragraph summary here]

## 6.3 Specific Recommendations

**Your Task:** Create a decision framework for algorithm implementation.

```{r recommendations-data}
# Create a summary table using your county reliability data
# Include: county name, median income, MOE percentage, reliability category

# Add a new column with algorithm recommendations using case_when():
# - High Confidence: "Safe for algorithmic decisions"
# - Moderate Confidence: "Use with caution - monitor outcomes"  
# - Low Confidence: "Requires manual review or additional data"

# Format as a professional table with kable()
```

**Key Recommendations:**

**Your Task:** Use your analysis results to provide specific guidance to the department.

1. **Counties suitable for immediate algorithmic implementation:** [List counties with high confidence data and explain why they're appropriate]

2. **Counties requiring additional oversight:** [List counties with moderate confidence data and describe what kind of monitoring would be needed]

3. **Counties needing alternative approaches:** [List counties with low confidence data and suggest specific alternatives - manual review, additional surveys, etc.]

## Questions for Further Investigation

[List 2-3 questions that your analysis raised that you'd like to explore further in future assignments. Consider questions about spatial patterns, time trends, or other demographic factors.]

# Technical Notes

**Data Sources:** 
- U.S. Census Bureau, American Community Survey 2018-2022 5-Year Estimates
- Retrieved via tidycensus R package on [date]

**Reproducibility:** 
- All analysis conducted in R version [your version]
- Census API key required for replication
- Complete code and documentation available at: [your portfolio URL]

**Methodology Notes:**
[Describe any decisions you made about data processing, county selection, or analytical choices that might affect reproducibility]

**Limitations:**
[Note any limitations in your analysis - sample size issues, geographic scope, temporal factors, etc.]

---

## Submission Checklist

Before submitting your portfolio link on Canvas:

- [ ] All code chunks run without errors
- [ ] All "[Fill this in]" prompts have been completed
- [ ] Tables are properly formatted and readable
- [ ] Executive summary addresses all four required components
- [ ] Portfolio navigation includes this assignment
- [ ] Census API key is properly set 
- [ ] Document renders correctly to HTML

**Remember:** Submit your portfolio URL on Canvas, not the file itself. Your assignment should be accessible at `your-portfolio-url/assignments/assignment_1/your_file_name.html`