---
title: "Assignment 2: Spatial Analysis and Visualization"
subtitle: "Healthcare Access and Equity in Pennsylvania"
author: "Christine"
date: October 10, 2025
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
---

## Assignment Overview

**Learning Objectives:**
- Apply spatial operations to answer policy-relevant research questions
- Integrate census demographic data with spatial analysis
- Create publication-quality visualizations and maps
- Work with spatial data from multiple sources
- Communicate findings effectively for policy audiences

---

## Part 1: Healthcare Access for Vulnerable Populations

### Research Question

**Which Pennsylvania counties have the highest proportion of vulnerable populations (elderly + low-income) living far from hospitals?**

Your analysis should identify counties that should be priorities for healthcare investment and policy intervention.

### Required Analysis Steps

Complete the following analysis, documenting each step with code and brief explanations:

#### Step 1: Data Collection (5 points)

Load the required spatial data:
- Pennsylvania county boundaries
- Pennsylvania hospitals (from lecture data)
- Pennsylvania census tracts

**Your Task:**
```{r}
# Load required packages
library(sf)
library(tidyverse)
library(tigris)
library(tidycensus)
library(scales)
library(patchwork)
library(here)
census_api_key("20068788c6e79d5716fbceb0dcd562ab23f74ca1")

# Load spatial data
# 1. Pennsylvania county boundaries
pa_counties <- st_read(here("labs/lab_2/data/Pennsylvania_County_Boundaries.shp"))

# 2. Pennsylvania hospitals (from lecture data)
hospitals <- st_read(here("labs/lab_2/data/hospitals.geojson"))

# 3. Pennsylvania census tracts
census_tracts <- tracts(state = "PA", cb = TRUE)

# Check that all data loaded correctly
```
```{r}
#check the CRS
cat("1. PA Counties:", st_crs(pa_counties)$input, "\n")
cat("2. Hospitals:", st_crs(hospitals)$input, "\n")
cat("3. Census Tracts:", st_crs(census_tracts)$input, "\n")
```

**Questions to answer:**
- How many hospitals are in your dataset? 
  - 223
- How many census tracts?
  - 3445
- What coordinate reference system is each dataset in? 
  - PA county boundaries CRS: PCS WGS 84 / Pseudo-Mercator
  - PA hospitals CRS: GCS WGS 84
  - PA census tracts CRS: GCS NAD 83

---

#### Step 2: Get Demographic Data 

Use `tidycensus` to download tract-level demographic data for Pennsylvania.

**Required variables:**
- Total population
- Median household income
- Population 65 years and over (you may need to sum multiple age categories)

**Your Task:**
```{r}
# Get demographic data from ACS
acs_vars <- c(
  total_pop = "B01003_001",
  med_hh_income = "B19013_001",
  # Male 65 years and over (sum of 6 categories)
  male_65_66 = "B01001_020",
  male_67_69 = "B01001_021",
  male_70_74 = "B01001_022",
  male_75_79 = "B01001_023",
  male_80_84 = "B01001_024",
  male_85_up = "B01001_025",
  # Female 65 years and over (sum of 6 categories)
  female_65_66 = "B01001_044",
  female_67_69 = "B01001_045",
  female_70_74 = "B01001_046",
  female_75_79 = "B01001_047",
  female_80_84 = "B01001_048",
  female_85_up = "B01001_049"
)

pa_demographics <- get_acs(
  geography = "tract",
  state = "PA",
  variables = acs_vars,
  output = "wide",
  year = 2022, 
  geometry = FALSE 
) %>%
  # Calculate the combined 65+ population and rename variables
  mutate(
    pop_65_over = male_65_66E + male_67_69E + male_70_74E + male_75_79E + male_80_84E + male_85_upE +
                  female_65_66E + female_67_69E + female_70_74E + female_75_79E + female_80_84E + female_85_upE
  ) %>%
  # Select and rename final columns for clarity and consistency
  select(
    GEOID,
    total_pop = total_popE,
    med_hh_income = med_hh_incomeE,
    pop_65_over
  )
# Join to tract boundaries
pa_tracts_demographics <- census_tracts %>%
 left_join(pa_demographics, by = "GEOID")

```

```{r}
#tracts have missing income data
missing_income_count <- pa_tracts_demographics %>%
  summarise(missing_count = sum(is.na(med_hh_income))) %>%
  pull(missing_count)
#median income across all PA census tracts
median_tract_income <- pa_tracts_demographics %>%
  filter(!is.na(med_hh_income)) %>%
  summarise(median_income = median(med_hh_income)) %>%
  pull(median_income)

```

**Questions to answer:**
- What year of ACS data are you using?
  - 2022 ACS data
- How many tracts have missing income data?
  - 62 tracts
- What is the median income across all PA census tracts?
  - $70,188
---

#### Step 3: Define Vulnerable Populations 

Identify census tracts with vulnerable populations based on TWO criteria:
1. Low median household income (choose an appropriate threshold)
2. Significant elderly population (choose an appropriate threshold)

**Your Task:**
```{r}
# Filter for vulnerable tracts based on your criteria
pa_tracts_demographics <- pa_tracts_demographics %>%
  mutate(pct_65_over = (pop_65_over / total_pop) * 100)

income_threshold <- quantile(pa_tracts_demographics$med_hh_income, 0.25, na.rm = TRUE)
elderly_threshold <- quantile(pa_tracts_demographics$pct_65_over, 0.75, na.rm = TRUE)

pa_tracts_demographics <- pa_tracts_demographics %>%
  mutate(
    income_low = med_hh_income < income_threshold,
    elderly_high = pct_65_over > elderly_threshold,
    vulnerable = if_else(income_low & elderly_high, TRUE, FALSE)
  )

```

```{r}
# Summary: how many tracts are vulnerable?
vulnerable_summary <- pa_tracts_demographics %>%
  summarise(
    total_tracts = n(),
    vulnerable_tracts = sum(vulnerable, na.rm = TRUE),
    pct_vulnerable = round((vulnerable_tracts / total_tracts) * 100, 1)
  )

vulnerable_summary
```

**Questions to answer:**
- What income threshold did you choose and why?
  - I defined low-income tracts as those with a median household income below the 25th percentile of all Pennsylvania census tracts. This data-driven threshold identifies the bottom quarter of tracts in terms of income, representing communities that are relatively economically disadvantaged compared to the rest of the state. Using a percentile-based cutoff avoids arbitrary dollar amounts and allows for consistent comparison across regions and time periods.
- What elderly population threshold did you choose and why?
  - I defined tracts with a significant elderly population as those where the percentage of residents aged 65 and over exceeds the 75th percentile statewide. This threshold highlights the top quartile of tracts with the highest concentration of older adults—populations that are more likely to experience increased healthcare needs and mobility challenges. By focusing on the upper quartile, the analysis targets areas where aging demographics are most pronounced and where healthcare accessibility is especially critical.
- How many tracts meet your vulnerability criteria?
  - 165 tracts
- What percentage of PA census tracts are considered vulnerable by your definition?
  - 4.8%
---

#### Step 4: Calculate Distance to Hospitals 

For each vulnerable tract, calculate the distance to the nearest hospital.

**Your Task:**
```{r}
# Transform to appropriate projected CRS
pa_tracts_proj <- st_transform(pa_tracts_demographics, crs = 5070)
hospitals_proj <- st_transform(hospitals, crs = 5070)

# Calculate distance from each tract centroid to nearest hospital 中心点 (Centroid)
tract_centroids <- st_centroid(pa_tracts_proj)

nearest_dist_m <- st_distance(tract_centroids, hospitals_proj) %>%
  apply(1, min)  # minimum distance for each tract
pa_tracts_proj <- pa_tracts_proj %>%
  mutate(
    dist_to_hospital_m = nearest_dist_m,
    dist_to_hospital_mi = dist_to_hospital_m / 1609.34
  )

vulnerable_distance_summary <- pa_tracts_proj %>%
  filter(vulnerable == TRUE) %>%
  summarise(
    avg_distance_mi = mean(dist_to_hospital_mi, na.rm = TRUE),
    max_distance_mi = max(dist_to_hospital_mi, na.rm = TRUE),
    over_15_miles = sum(dist_to_hospital_mi > 15, na.rm = TRUE)
  )

vulnerable_distance_summary

```

**Requirements:**
- Use an appropriate projected coordinate system for Pennsylvania
- Calculate distances in miles
- Explain why you chose your projection

**Questions to answer:**
- What is the average distance to the nearest hospital for vulnerable tracts?
  - 4.75 miles
- What is the maximum distance?
  - 19.32 miles
- How many vulnerable tracts are more than 15 miles from the nearest hospital?
  - 10 tracts
- Why you chose your projection
  - For this analysis, I used the NAD83 / Conus Albers Equal Area projection (EPSG:5070). This projection is commonly used for spatial analysis across the continental United States because it preserves both area and distance reasonably well over large regions. Pennsylvania spans multiple UTM zones, so using a single statewide or national projection like Albers Equal Area helps avoid spatial distortion that would occur near zone boundaries. The units of this projection are in meters, which allows for accurate distance calculations when measuring the proximity of census tracts to hospitals. Overall, EPSG:5070 provides a balance between accuracy and consistency, making it appropriate for statewide healthcare accessibility analysis.

---

#### Step 5: Identify Underserved Areas 

Define "underserved" as vulnerable tracts that are more than 15 miles from the nearest hospital.

**Your Task:**
```{r}
# Create underserved variable
pa_tracts_proj <- pa_tracts_proj %>%
  mutate(
    underserved = if_else(vulnerable == TRUE & dist_to_hospital_mi > 15, TRUE, FALSE)
  )
# Summarize underserved tracts
underserved_summary <- pa_tracts_proj %>%
  summarise(
    total_tracts = n(),
    vulnerable_tracts = sum(vulnerable, na.rm = TRUE),
    underserved_tracts = sum(underserved, na.rm = TRUE),
    pct_underserved = round((underserved_tracts / vulnerable_tracts) * 100, 1)
  )

underserved_summary
```

**Questions to answer:**
- How many tracts are underserved?
  - 10 tracts
- What percentage of vulnerable tracts are underserved?
  - 6.1%
- Does this surprise you? Why or why not?
  - The results show that only a small percentage of vulnerable tracts are located more than 15 miles from the nearest hospital. This is not particularly surprising, as Pennsylvania has a relatively dense healthcare network, especially around urban and suburban areas. However, the underserved tracts that do exist are likely concentrated in rural or mountainous regions where population density is low and travel distances are longer. These areas may require targeted healthcare investment or mobile medical services to improve access.

---

#### Step 6: Aggregate to County Level

Use spatial joins and aggregation to calculate county-level statistics about vulnerable populations and hospital access.

**Your Task:**
```{r}
# Spatial join tracts to counties
pa_counties_proj <- st_transform(pa_counties, st_crs(pa_tracts_proj))
tracts_county_join <- st_join(pa_tracts_proj, pa_counties_proj, join = st_intersects, left = TRUE)

# Aggregate statistics by county
county_summary <- tracts_county_join %>%
  st_drop_geometry() %>%
  group_by(COUNTY_NAM) %>%
  summarise(
    n_tracts = n(),
    vulnerable_tracts = sum(vulnerable, na.rm = TRUE),
    underserved_tracts = sum(underserved, na.rm = TRUE),
    pct_underserved_vulnerable = round((underserved_tracts / vulnerable_tracts) * 100, 1),
    avg_distance_vulnerable = mean(dist_to_hospital_mi[vulnerable == TRUE], na.rm = TRUE),
    total_vulnerable_pop = sum(total_pop[vulnerable == TRUE], na.rm = TRUE),
    underserved_vulnerable_pop = sum(total_pop[vulnerable == TRUE & underserved == TRUE], na.rm = TRUE)
  ) %>%
  arrange(desc(pct_underserved_vulnerable))

head(county_summary, 10)
```
```{r}
# Plot map of percent underserved by Pennsylvania counties.

county_map <- pa_counties_proj %>%
  left_join(county_summary %>% select(COUNTY_NAM, pct_underserved_vulnerable),
            by = c("COUNTY_NAM" = "COUNTY_NAM"))

ggplot(county_map) +
  geom_sf(aes(fill = pct_underserved_vulnerable), color = "white", size = 0.5) +
  scale_fill_viridis_c(
    name = "Percent Underserved",
    option = "magma",
    # Format to show percent sign on color bar.
    labels = function(x) paste0(x, "%")
  ) +
  labs(
    title = "Most Underserved Counties",
    subtitle = "Pennsylvania, United States",
    caption = "Source: ACS 2019–2023"
  ) +
  theme_void()
```

**Required county-level statistics:**
- Number of vulnerable tracts
- Number of underserved tracts  
- Percentage of vulnerable tracts that are underserved
- Average distance to nearest hospital for vulnerable tracts
- Total vulnerable population

```{r}
top5_pct <- county_summary %>% 
  arrange(desc(pct_underserved_vulnerable)) %>% 
  slice_head(n = 5)
top5_pct

top1_pop <- county_summary %>% 
  arrange(desc(underserved_vulnerable_pop)) %>% 
  slice_head(n = 1)
top1_pop

```

**Questions to answer:**
- Which 5 counties have the highest percentage of underserved vulnerable tracts?
  - The top five counties with the highest percentage of underserved vulnerable tracts are BRADFORD,COLUMBIA,JUNIATA,MONROE and PERRY. They are primarily rural, located in northern and central Pennsylvania. These areas tend to have sparse hospital coverage and lower population density, contributing to longer travel distances for healthcare access.
  
- Which counties have the most vulnerable people living far from hospitals?
  - PERRY has the largest underserved vulnerable population, reflecting both a relatively high overall share of vulnerable residents and the presence of several tracts located far from hospitals. This suggests that limited healthcare infrastructure and rural settlement patterns contribute to significant access challenges for vulnerable groups in the county.
  
- Are there any patterns in where underserved counties are located?
  - Underserved counties are primarily located in rural and mountainous regions of Pennsylvania, particularly in the north-central and western parts of the state. These areas have lower population density, fewer major healthcare facilities, and longer travel distances between communities, making them more susceptible to healthcare access inequities.
---

#### Step 7: Create Summary Table 

Create a professional table showing the top 10 priority counties for healthcare investment.

**Your Task:**
```{r}
# Create and format priority counties table
library(knitr)
library(scales)

# 7.1 Select and format key variables for display
priority_counties <- county_summary %>%
  arrange(desc(pct_underserved_vulnerable)) %>%    
  slice_head(n = 10) %>%    
  select(
    County = COUNTY_NAM,
    `Vulnerable Tracts` = vulnerable_tracts,
    `Underserved Tracts` = underserved_tracts,
    `% Underserved` = pct_underserved_vulnerable,
    `Avg Distance (mi)` = avg_distance_vulnerable,
    `Total Vulnerable Population` = total_vulnerable_pop
  ) %>%
  mutate(
    `Total Vulnerable Population` = comma(`Total Vulnerable Population`),  # add commas
    `% Underserved` = percent(`% Underserved` / 100, accuracy = 0.1),
    `Avg Distance (mi)` = round(`Avg Distance (mi)`, 1)
  )

# 2. Create professional summary table
kable(
  priority_counties,
  caption = "Table 1. Top 10 Pennsylvania Counties with the Highest Percentage of Underserved Vulnerable Tracts",
  align = "lccccc",
  digits = 1,
  format = "html"
)

```

**Requirements:**
- Use `knitr::kable()` or similar for formatting
- Include descriptive column names
- Format numbers appropriately (commas for population, percentages, etc.)
- Add an informative caption
- Sort by priority (you decide the metric)

---

## Part 2: Comprehensive Visualization 

Using the skills from Week 3 (Data Visualization), create publication-quality maps and charts.

### Map 1: County-Level Choropleth 

Create a choropleth map showing healthcare access challenges at the county level.

**Your Task:**
```{r}
# Create county-level access map
library(ggplot2)
library(viridis)

county_map_data <- pa_counties_proj %>%
  left_join(county_summary, by = c("COUNTY_NAM" = "COUNTY_NAM"))

ggplot() +
  # Counties filled by % underserved vulnerable tracts
  geom_sf(data = county_map_data, aes(fill = pct_underserved_vulnerable), color = "white", size = 0.3) +
  
  # Hospital locations as points
  geom_sf(data = hospitals_proj, color = "red", size = 1, alpha = 0.7) +
  
  # Color scale for access challenge
  scale_fill_viridis(
    name = "% Underserved Vulnerable Tracts",
    option = "C",
    direction = -1,
    labels = function(x) paste0(x, "%")
  ) +
  
  # Map titles and caption
  labs(
    title = "Healthcare Access Challenges Across Pennsylvania Counties",
    subtitle = "Counties shaded by the percentage of vulnerable tracts located more than 15 miles from the nearest hospital",
    caption = "Data sources: American Community Survey (2022), Pennsylvania Hospital Data, and U.S. Census TIGER/Line Shapefiles"
  ) +
  
  # Clean minimal theme
  theme_void() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 10),
    plot.caption = element_text(size = 9, color = "gray40"),
    legend.position = "bottom"
  )
```

**Requirements:**
- Fill counties by percentage of vulnerable tracts that are underserved
- Include hospital locations as points
- Use an appropriate color scheme
- Include clear title, subtitle, and caption
- Use `theme_void()` or similar clean theme
- Add a legend with formatted labels

---

### Map 2: Detailed Vulnerability Map 

Create a map highlighting underserved vulnerable tracts.

**Your Task:**
```{r}
# Create detailed tract-level map
ggplot() +
  # 1. 所有普查区背景（灰色）——表示非弱势区域
  geom_sf(data = pa_tracts_proj, fill = "grey90", color = NA) +
  
  # 2. 弱势但不服务不足的区域（浅黄色）
  geom_sf(data = filter(pa_tracts_proj, vulnerable == TRUE & underserved == FALSE),
          fill = "#FED976", color = NA) +
  
  # 3. 服务不足的弱势区域（亮红色，最突出）
  geom_sf(data = filter(pa_tracts_proj, underserved == TRUE & vulnerable == TRUE),
          fill = "#E31A1C", color = "white", size = 0.05) +
  
  # 4. 县界线（白色）
  geom_sf(data = pa_counties_proj, fill = NA, color = "white", size = 0.3) +
  
  # 5. 医院位置（蓝色点）
  geom_sf(data = hospitals_proj, color = "blue", size = 0.8, alpha = 0.7) +
  
  # 6. 添加标题、说明与数据来源
  labs(
    title = "Underserved Vulnerable Tracts in Pennsylvania",
    subtitle = "Red areas represent vulnerable tracts located more than 15 miles from the nearest hospital",
    caption = "Data sources: ACS (2022), Pennsylvania Hospital Data, U.S. Census TIGER/Line Shapefiles"
  ) +
  
  # 7. 设置样式
  theme_void() +
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 12),
    plot.caption = element_text(size = 9, color = "grey50")
  )



```

**Requirements:**
- Show underserved vulnerable tracts in a contrasting color
- Include county boundaries for context
- Show hospital locations
- Use appropriate visual hierarchy (what should stand out?)
- Include informative title and subtitle

---

### Chart: Distribution Analysis

Create a visualization showing the distribution of distances to hospitals for vulnerable populations.

**Your Task:**
```{r}
# Create distribution visualization
library(ggplot2)
library(scales)

vulnerable_dist <- pa_tracts_demographics %>%
  filter(vulnerable == TRUE)
ggplot(pa_tracts_proj, aes(x = dist_to_hospital_mi)) +
  geom_histogram(
    aes(y = ..density..), 
    bins = 30, 
    fill = "#74add1", 
    color = "white", 
    alpha = 0.8
  ) +
  geom_density(color = "#0571b0", size = 1) +
  geom_vline(xintercept = 15, linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = 15.5, y = 0.02, label = "15-mile threshold", hjust = 0, color = "red", size = 3) +
  labs(
    title = "Distribution of Distance to Nearest Hospital for Vulnerable Tracts",
    subtitle = "Dashed line indicates the 15-mile threshold used to define underserved areas",
    x = "Distance to Nearest Hospital (miles)",
    y = "Density",
    caption = "Source: ACS (2022), Pennsylvania Hospital Data, TIGER/Line Shapefiles"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 11),
    plot.caption = element_text(size = 9, color = "gray50")
  )
```

**Suggested chart types:**
- Histogram or density plot of distances
- Box plot comparing distances across regions
- Bar chart of underserved tracts by county
- Scatter plot of distance vs. vulnerable population size

**Requirements:**
- Clear axes labels with units
- Appropriate title
- Professional formatting
- Brief interpretation (1-2 sentences as a caption or in text)
  - **11111**
---

## Part 3: Bring Your Own Data Analysis 

Choose your own additional spatial dataset and conduct a supplementary analysis.

### Challenge Options

Choose ONE of the following challenge exercises, or propose your own research question using OpenDataPhilly data (https://opendataphilly.org/datasets/). 

**Note these are just loose suggestions to spark ideas - follow or make your own as the data permits and as your ideas evolve. This analysis should include bringing in your own dataset, ensuring the projection/CRS of your layers align and are appropriate for the analysis (not lat/long or geodetic coordinate systems). The analysis portion should include some combination of spatial and attribute operations to answer a relatively straightforward question**

---

#### Education & Youth Services

**Option A: Educational Desert Analysis**
- **Data:** Schools, Libraries, Recreation Centers, Census tracts (child population)
- **Question:** "Which neighborhoods lack adequate educational infrastructure for children?"
- **Operations:** Buffer schools/libraries (0.5 mile walking distance), identify coverage gaps, overlay with child population density
- **Policy relevance:** School district planning, library placement, after-school program siting

**Option B: School Safety Zones**
- **Data:** Schools, Crime Incidents, Bike Network
- **Question:** "Are school zones safe for walking/biking, or are they crime hotspots?"
- **Operations:** Buffer schools (1000ft safety zone), spatial join with crime incidents, assess bike infrastructure coverage
- **Policy relevance:** Safe Routes to School programs, crossing guard placement

---

#### Environmental Justice

**Option C: Green Space Equity** 
- **Data:** Parks, Street Trees, Census tracts (race/income demographics)
- **Question:** "Do low-income and minority neighborhoods have equitable access to green space?"
- **Operations:** Buffer parks (10-minute walk = 0.5 mile), calculate tree canopy or park acreage per capita, compare by demographics
- **Policy relevance:** Climate resilience, environmental justice, urban forestry investment
---

#### Public Safety & Justice

**Option D: Crime & Community Resources**
- **Data:** Crime Incidents, Recreation Centers, Libraries, Street Lights
- **Question:** "Are high-crime areas underserved by community resources?"
- **Operations:** Aggregate crime counts to census tracts or neighborhoods, count community resources per area, spatial correlation analysis
- **Policy relevance:** Community investment, violence prevention strategies
---

#### Infrastructure & Services

**Option E: Polling Place Accessibility**
- **Data:** Polling Places, SEPTA stops, Census tracts (elderly population, disability rates)
- **Question:** "Are polling places accessible for elderly and disabled voters?"
- **Operations:** Buffer polling places and transit stops, identify vulnerable populations, find areas lacking access
- **Policy relevance:** Voting rights, election infrastructure, ADA compliance

---

#### Health & Wellness

**Option F: Recreation & Population Health**
- **Data:** Recreation Centers, Playgrounds, Parks, Census tracts (demographics)
- **Question:** "Is lack of recreation access associated with vulnerable populations?"
- **Operations:** Calculate recreation facilities per capita by neighborhood, buffer facilities for walking access, overlay with demographic indicators
- **Policy relevance:** Public health investment, recreation programming, obesity prevention

---

#### Emergency Services

**Option G: EMS Response Coverage**
- **Data:** Fire Stations, EMS stations, Population density, High-rise buildings
- **Question:** "Are population-dense areas adequately covered by emergency services?"
- **Operations:** Create service area buffers (5-minute drive = ~2 miles), assess population coverage, identify gaps in high-density areas
- **Policy relevance:** Emergency preparedness, station siting decisions

---

#### Arts & Culture

**Option H: Cultural Asset Distribution**
- **Data:** Public Art, Museums, Historic sites/markers, Neighborhoods
- **Question:** "Do all neighborhoods have equitable access to cultural amenities?"
- **Operations:** Count cultural assets per neighborhood, normalize by population, compare distribution across demographic groups
- **Policy relevance:** Cultural equity, tourism, quality of life, neighborhood identity

---

### Data Sources

**OpenDataPhilly:** https://opendataphilly.org/datasets/
- Most datasets available as GeoJSON, Shapefile, or CSV with coordinates
- Always check the Metadata for a data dictionary of the fields.

**Additional Sources:**
- **Pennsylvania Open Data:** https://data.pa.gov/
- **Census Bureau (via tidycensus):** Demographics, economic indicators, commute patterns
- **TIGER/Line (via tigris):** Geographic boundaries

### Recommended Starting Points

**If you're feeling confident:** Choose an advanced challenge with multiple data layers. 
**If you are a beginner, choose something more manageable that helps you understand the basics**


**If you have a different idea:** Propose your own question! Just make sure:
- You can access the spatial data
- You can perform at least 2 spatial operations

### Your Analysis

**Your Task:**

1. **Find and load additional data**
   - Document your data source
   - Check and standardize the CRS
   - Provide basic summary statistics

```{r}
# Load your additional dataset




```

**Questions to answer:**
- What dataset did you choose and why?
- What is the data source and date?
- How many features does it contain?
- What CRS is it in? Did you need to transform it?

---

2. **Pose a research question**

Write a clear research statement that your analysis will answer.

**Examples:**
- "Do vulnerable tracts have adequate public transit access to hospitals?"
- "Are EMS stations appropriately located near vulnerable populations?"
- "Do areas with low vehicle access have worse hospital access?"

---

3. **Conduct spatial analysis**

Use at least TWO spatial operations to answer your research question.

**Required operations (choose 2+):**
- Buffers
- Spatial joins
- Spatial filtering with predicates
- Distance calculations
- Intersections or unions
- Point-in-polygon aggregation

**Your Task:**
```{r}
# Your spatial analysis










```

**Analysis requirements:**
- Clear code comments explaining each step
- Appropriate CRS transformations
- Summary statistics or counts
- At least one map showing your findings
- Brief interpretation of results (3-5 sentences)

**Your interpretation:**

[Write your findings here]


## Finally - A few comments about your incorporation of feedback!
Take a few moments to clean up your markdown document and then write a line or two or three about how you may have incorporated feedback that you recieved after your first assignment. 

---

## Submission Requirements

**What to submit:**

1. **Rendered HTML document posted to your course portfolio** with all code, outputs, maps, and text
   - Use `embed-resources: true` in YAML so it's a single file
   - All code should run without errors
   - All maps and charts should display correctly

**File naming:** `LastName_FirstName_Assignment2.html` and `LastName_FirstName_Assignment2.qmd`

---

